---
title: "Matching cells and donors (Cuomo 2020)"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---
*The aim of this notebook is to map cells to donors based on the available meta-data and recommendations from Anna Cuomo in our email correspondence*

Load packages
```{r setup}
library(data.table)
library(magrittr)
library(tidyr)
library(dplyr)
```
Load the 2 datasets
```{r}
ena.runs.sc <- fread("/Users/lakshmichanemougam/Desktop/Battle_Lab/getting_fastq_files/filereport_read_run_PRJEB14362_tsv.txt") %>% print()
zenodo.meta <- fread("/Users/lakshmichanemougam/Desktop/Battle_Lab/cuomo_supplementary/cell_metadata_cols.tsv") %>% print()
#Make sure these are the same
stopifnot(all(zenodo.meta$V1 == zenodo.meta$cell_name))
```

Extract the cell identifiers from the CRAM files:
```{r}
ena.runs.sc <- ena.runs.sc %>% mutate("cell_name"=ena.runs.sc$submitted_ftp %>% basename(.) %>% gsub(pattern = ".cram", replacement = "", x = .))
head(ena.runs.sc$cell_name)
```
How many unique cell ids appear in each file?
```{r}
length(unique(ena.runs.sc$cell_name))
length(unique(zenodo.meta$cell_name))
```
*There are more cells in the ENA runs- these may be ones that got dropped for QC perhaps?*
How many ENA runs are there for which we don't have Zenodo data available (the inverse of the above)?
```{r}
length(unique(ena.runs.sc$cell_name))
length(unique((ena.runs.sc %>% filter(!(cell_name %in% zenodo.meta$cell_name)))$cell_name))
```
*It looks like only about half of all the cell_ids correspond with a cell id used in the downstream analysis (in the meta-analysis zenodo file). These runs omitted in the downstream analysis may have been dropped for QC or other reasons. Might want to follow-up.*

Select only the zenodo meta-data entries for which we have ENA runs available:
```{r}
short.list <- zenodo.meta %>% filter(cell_name %in% ena.runs.sc$cell_name) %>% print()
length(unique(short.list$cell_name))
```
*So it looks like only 23793 of the cell_ids from Zenodo are in ENA.*

## Join the datasets: find donor ids corresponding to each cell
Create a scaled down version of the ENA data to merge with the zenodo data
```{r}
merge.list <- ena.runs.sc %>% filter(cell_name %in% short.list$cell_name) %>% select(cell_name, study_accession, sample_accession, submitted_ftp, fastq_ftp) 
#Join the Zenodo and ENA data by the cellIDs, mapping donor ids onto the individual runs
ena.zenodo.join <- merge(short.list,merge.list, by = "cell_name" )
#save this file for downstream use
#write.csv(ena.zenodo.join, file ="/Users/lakshmichanemougam/Downloads/merge_ena_zenodo.csv" , quote = FALSE, row.names = FALSE)
#visualize this
ena.zenodo.join %>% select(cell_name, donor_long_id, submitted_ftp)
nrow(ena.zenodo.join)
length(unique(ena.zenodo.join$donor_long_id))
```
*this means we have cell results for 103 distinct donors, across a total of 23793 cells. AWESOME!*

## Identify how many cell lines/ donors missing from ENA
Anna mentioned that "a small subset of lines (~10%) will not be in included on ENA as those are diseased cell lines" and so require managed access.
I assume this means about 10% of donors in the final Zenodo file will have no cell data available on ENA
```{r}
missing.samples <- unique(zenodo.meta$donor)[!(unique(zenodo.meta$donor) %in% unique(ena.zenodo.join$donor))]
message("Percent missing: ", length(missing.samples)/length(unique(zenodo.meta$donor)) * 100)

```
This is close enough.

**Do these donors explain all of the difference in cell_ids between the Zenodo a ENA data?**
```{r}
donors.with.any.missing.cells <- unique((zenodo.meta %>% filter(!(cell_name %in% ena.runs.sc$cell_name)))$donor_long_id)
donors.with.any.missing.cells[!(donors.with.any.missing.cells %in% missing.samples)]
```
*These donors have cell data referenced in Zenodo that is not available in ENA. Strange*

## Some considerations moving forward:
 - Can we explain why half of the ENA cells aren't included in the Zenodo data?
 - Can we explain why so many cells included in the Zenodo data and associated with non-diseased lines are omitted?
 - For our purpose of testing genotype calling and imputation, I recommend we select an individual for whom we know we have all of the cell data (so not someone on the list above)
