---
title: "ena_zenodo_genotype"
output: html_document
---
*This compares the merged zenodo and ENA csv with list of available genotypes*
```{r}
#To run on rockfish, always add personal library
.libPaths(c(.libPaths(), "/home/lchanem1/rlibs/4.0.2/gcc/9.3.0"))
```

Load packages
```{r setup}
library(data.table)
library(magrittr)
library(tidyr)
library(dplyr)
library(stringr, lib.loc = "/home/lchanem1/rlibs/4.0.2/gcc/9.3.0")
```

Loading datasets
```{r}
ena.genotype <- fread("filereport_analysis_PRJEB11750_tsv.txt") %>% print()
ena.report <- fread("filereport_read_run_PRJEB14362_tsv.txt") %>% print()
ena.zenodo.merged <- fread("merge_ena_zenodo.csv") %>% print()
```

Extract donor long IDs from .cram filenames
```{r}
ena.genotype$donor_ids <- str_extract(ena.genotype$submitted_ftp, "(?<=\\/)[A-Za-z0-9_-]+(?=\\.)")
donor_ids <- as.array(ena.genotype$donor_ids)
donor_long_id <- as.array(ena.zenodo.merged$donor_long_id)
```

Comparing the donor_long_ids on ENA Genotype and 'pseudo' on merged_ena_zenodo
```{r}
head(donor_ids)
head(donor_long_id)
length(intersect(donor_ids, donor_long_id))
```

Filter ena.zenodo.merged for only those donor_long_ids that are present in donor_ids of ena.genotype
```{r}
merged.final <- ena.zenodo.merged[ena.zenodo.merged$donor_long_id %in% donor_ids]
head(merged.final)
write.csv(merged.final, "merged_final.csv", quote = FALSE, row.names = FALSE)
```

How many donors having genotype information published?
```{r}
length(unique(ena.zenodo.merged$donor_long_id))
length(unique(merged.final$donor_long_id))
```

Finding number of cells per donors in the final doc:
```{r}
total_donors <- as.array(merged.final$donor_long_id)
freq <- table(total_donors)
freq <- freq[order(freq)]
freq
```

Final comment: Out of 104 donors (further down from 125 original donors as claimed), 89 have associated ENA cram files, metadata and genotype information.

Highest number of cells belongs to donor long ID "HPSI0114i-joxm_1" with 1328 cells.

## Plot all donors (all 89) and their cell numbers in ascending order
```{r}
freq_df <- data.frame(donor = names(freq), frequency = as.numeric(freq))
barplot(freq_df$frequency, names.arg = freq_df$donor, xlab = "Donor Names", ylab = "No. of Cells", main = "Donor Cell Count")
axis(side = 1, at = highlight_indices, labels = FALSE, tick = TRUE)
keywords <- c("HPSI0115i-zihe_1","HPSI0114i-iisa_3","HPSI0614i-uilk_3","HPSI0414i-xojn_3","HPSI0914i-qehq_3","HPSI0114i-eipl_1","HPSI0514i-letw_1","HPSI0314i-sojd_3","HPSI0214i-poih_4","HPSI0114i-joxm_1")
highlight_indices <- which(freq_df$donor %in% keywords)
text(highlight_indices, freq_df$frequency[highlight_indices], labels = freq_df$donor[highlight_indices], pos = 3, cex = 0.8, col = "red")

```

###Downloading all Cuomo data for donors with available genotype (out of 89 donors, 10 already downloaded; remaining 79 preprocessing)

```{r}
keywords <- c("HPSI0115i-zihe_1","HPSI0114i-iisa_3","HPSI0614i-uilk_3","HPSI0414i-xojn_3","HPSI0914i-qehq_3","HPSI0114i-eipl_1","HPSI0514i-letw_1","HPSI0314i-sojd_3","HPSI0214i-poih_4","HPSI0114i-joxm_1")

# Subset data to include all donors except for 10 donors in keywords
merged.final <- data.frame(merged.final)
remaining <- dplyr::filter(merged.final, !donor_long_id %in% keywords)
donors_list <- unique(remaining$donor_long_id)

# Get list of fastq ftps for each donor in donors_list
for (donor_name in donors_list) {
  temp <- remaining[remaining$donor_long_id %in% donor_name, ]
  for (i in 1:nrow(temp)) {
    file_name <- paste0(temp[i, "donor_short_id"], "_fastq_ftp.txt")
    write.table(temp[i, "fastq_ftp", drop = FALSE], file = file_name, row.names = FALSE, col.names = FALSE, quote = FALSE, append = TRUE)
    alter_ftp <- readLines(file_name)
    alter_ftp <- gsub(";","\n", alter_ftp)
    writeLines(alter_ftp, file_name)
  } 
}

## Get genotype file ftps for all donor in donors_list
for (donor_name in donors_list){
  temp <- ena.genotype[ena.genotype$donor_ids %in% donor_name, ]
  for (i in 1:nrow(temp)) {
    write.table(temp[i, "submitted_ftp", drop = FALSE], file = "remaining_genotypes_ftp.txt", row.names = FALSE, col.names = FALSE, quote = FALSE, append = TRUE)
    alter_ftp <- readLines("remaining_genotypes_ftp.txt")
    alter_ftp <- gsub(";","\n", alter_ftp)
    writeLines(alter_ftp, "remaining_genotypes_ftp.txt")
  }
}
```


------------------------------------------FOR SAMPLE 10 DONORS ALONE-------------------------------------------
### Creating a dataframe with just the final choice of donors, and generating .txt file with ftps of all donors
```{r}
keywords <- c("HPSI0115i-zihe_1","HPSI0114i-iisa_3","HPSI0614i-uilk_3","HPSI0414i-xojn_3","HPSI0914i-qehq_3","HPSI0114i-eipl_1","HPSI0514i-letw_1","HPSI0314i-sojd_3","HPSI0214i-poih_4","HPSI0114i-joxm_1")
gencove_metadata <- merged.final[merged.final$donor_long_id %in% keywords]
gencove_metadata <- select(gencove_metadata, donor_short_id, donor_long_id, fastq_ftp, submitted_ftp)
metadata_md5 <- merge(gencove_metadata, ena.report, by="fastq_ftp")
metadata_md5$submitted_md5 <- sapply(strsplit(as.character(metadata_md5$submitted_md5), ";"), `[`, 1)
```

```{r}
# Get .cram files for each of the 10 donors
for (keyword in keywords) {
  temp <- gencove_metadata[gencove_metadata$donor_long_id %in% keyword, ]
  for (i in 1:nrow(temp)) {
    file_name <- paste0(temp[i, "donor_short_id"], "_cram_ftp.txt")
    write.table(temp[i, "submitted_ftp", drop = FALSE], file = file_name, row.names = FALSE, col.names = FALSE, quote = FALSE, append = TRUE)
    alter_ftp <- readLines(file_name)
    alter_ftp <- gsub("#","%23", alter_ftp) ## wget links aren't correct and # needs to be replaced with %23
    alter_ftp <- gsub("http:", "ftp:", alter_ftp) ## corrected links
    writeLines(alter_ftp, file_name)
  } 
}

## Get md5 for all downloaded crams
write.table(metadata_md5$submitted_md5, file = "all_crams_md5.txt", row.names = FALSE, col.names = FALSE, quote = FALSE, append = TRUE)
```

### Getting a txt file with all genotype ftps associated with chosen final donors
```{r}
for (keyword in keywords){
  temp <- ena.genotype[ena.genotype$donor_ids %in% keyword, ]
  for (i in 1:nrow(temp)) {
    write.table(temp[i, "submitted_ftp", drop = FALSE], file = "all_genotypes_ftp.txt", row.names = FALSE, col.names = FALSE, quote = FALSE, append = TRUE)
    alter_ftp <- readLines("all_genotypes_ftp.txt")
    alter_ftp <- gsub(";","\n", alter_ftp)
    writeLines(alter_ftp, "all_genotypes_ftp.txt")
  }
}
```

### NOT NEEDED ANYMORE: Final Gencove metadata file formatting including fastq pathnames, client_id, r_notation
```{r}
gencove_metadata <- separate_rows(gencove_metadata, fastq_ftp, sep = ";")
gencove_metadata <- mutate(gencove_metadata, r_notation = "")
gencove_metadata <- select(gencove_metadata, donor_long_id, r_notation, fastq_ftp)
gencove_metadata <- rename(gencove_metadata, client_id = donor_long_id, path = fastq_ftp)
head(gencove_metadata)
```

```{r}
gencove_metadata$path <- substring(gencove_metadata$path, nchar(gencove_metadata$path)-20, nchar(gencove_metadata$path))
gencove_metadata$path <- paste0("/home/lchanem1/data-abattle4/lakshmi/cuomo_2020/raw_reads/", gencove_metadata$path)
gencove_metadata$r_notation <- ifelse(substr(gencove_metadata$path, nchar(gencove_metadata$path)-9, nchar(gencove_metadata$path)-9) == "1", "r1", "r2")
gencove_metadata$client_id <- gsub("_","-", gencove_metadata$client_id)
head(gencove_metadata)
write.csv(gencove_metadata, "gencove_metadata.fastq-map.csv", quote = FALSE, row.names = FALSE)
```
